# R Language Control Structures {#control-structures}

In chapter [Functions] we briefly described conditional execution
using the `if`/`else`
construct.  Here we introduce the other main control structures, and
cover the if/else statement in more detail.


## Loops

One of the common problems that computers are good at, is to execute
similar tasks repeatedly many times.  This can be achieved through
_loops_.  R implements three loops: for-loop, while-loop, and
repeat-loop, these are among the most popular loops in many modern
programming languages.  Loops are often advised agains in case of R.  While it
is true that loops in R are more demanding than in several other
languages, and can often be replaced by more efficient constructs,
they nevertheless have a very important role in R.  Good usage of loops is
easy, fast, and improves the readability of your code.


### For Loop

For-loop is a good choice if you want to run some statements
repeatedly, each time picking a different value for a variable from a
sequence;
or just repeat some
statements a given number of times.  The syntax of for-loop is the
following:
```r
for(variable in sequence) {
                           # do something many times.  Each time 'variable' has
                           # a different value, taken from 'sequence'
}
```
The statement block inside of the curly braces {} is
repeated through the sequence.

Despite the curly braces, the loop body is still evaluated in the same
environment as the rest of the code.  This means the variables
generated before entering the loop are visible from inside, and the
variables created inside will be visible after the loop ends.  This is
in contrast to functions where the function body, also enclosed in curly
braces, is a separate scope.

A simple and perhaps the most common usage of for loop is just to run
some code a given number of times:
```{r}
for(i in 1:5)
   print(i)
```
Through the loop the variable `i` takes all the values from the
sequence `1:5`.  Note: we do not need curly braces if the loop block
only contains a single statement.

But the sequence does not have to be a simple sequence of numbers.  It
may as well be a list:
```{r}
data <- list("Xiaotian", 25, c(2061234567, 2069876543))
for(field in data) {
   cat("field length:", length(field), "\n")
}
```

As another less common example, we may run different functions over
the same data:
```{r}
data <- 1:4
for(fun in c(sqrt, exp, sin)) {
   print(fun(data))
}
```

### While-loop

While-loop let's the programmer to decide explicitly if we want to execute the
statements one more time:
```r
while(condition) {
                           # do something if the 'condition' is true
}
```

For instance, we can repeat something a given number of times
(although this is more of a task for for-loop):
```{r}
i <- 0
while(i < 5) {
   print(i)
   i <- i + 1
}
```

Note that in case the condition is not true to begin with, the loop never
executes and control is immediately passed to the code after the loop:
```{r}
i <- 10
while(i < 5) {
   print(i)
   i <- i + 1
}
cat("done\n")
```

While-loop is great for handling repeated attempts to get input in
shape.  One can write something like this:
```r
data <- getData()
while(!dataIsGood(data)) {
   data <- getData(tryReallyHard = TRUE)
}
```

This will run `getData()` until data is good.  If it is not good at
the first try, the code tries really hard next time(s).  

While can
also easily do infinite loops with:
```r
while(TRUE) { ... }
```
will execute until something breaks the loop.  Another somewhat
interesting usage of `while` is to disable execution of part of your
code:
```r
while(FALSE) {
   ## code you don't want to be executed
}
```


### repeat-loop

The R way of implementing repeat-loop does not include any condition
and hence it just keeps running infdefinitely, or until something breaks
the execution (see below):
```r
repeat {
   ## just do something indefinitely many times.
   ## if you want to stop, use `break` statement.
}
```

Repeat is useful in many similar situations as while.  For instance,
we can write a modified version of getting data using `repeat`:
```r
repeat {
   data <- getData()
   if(dataIsGood())
      break
}
```

The main difference between this, and the previous `while`-version is
that we a) don't have to write `getData` function twice, and b) on
each try we attempt to get data in the same way.

Exactly as in the case of _for_ and _while_ loop, enclosing the loop body in curly braces
is only necessary if it contains more than one statement.  For instance, if you want to close all the graphics
devices left open by buggy code, you can use the following one-liner
from console:
```r
repeat dev.off()
```
The loop will be broken by error when attempting to
close the last null device.

It should also be noted that the loop does not create a separate
environment (scope).  All the variables from outside of the loop are
visible inside, and variables created inside of the loop remain
accessible after the loop terminates.


### Leaving Early: `break` and `next`

A straightforward way to leave a loop is the `break` statement.  It just leaves
the loop and transfers control to the code immediately after the loop
body.  It breaks all three loops discussed here (_for_, _while_, and _repeat_-loops)
but not some of the other types of loops implemented outside of base R.  For instance:
```{r}
for(i in 1:10) {
   cat(i, "\n")
   if(i > 4)
      break
}
cat("done\n")
```

As the loop body will use the same environment (the same variable
values) as the rest of the code, the loop variables can be used right
afterwards.  For instance, we can see that `i = 5`:
```{r}
print(i)
```

Opposite to `break`, `next` will throw the execution flow back to the
head of the loop without running the commands following `next`.  In
case of for-loop, this means a new value is picked from the sequence,
in case of while-loop the condition is evaluated again; the head of
repeat loop does not do any calculations.  For instance, we can print
only even numbers:
```{r}
for(i in 1:10) {
   if(i %% 2 != 0)
      next
   print(i)
}
```
`next` jumps over printing where modulo of division by 2 is not 0.


### When (Not) To Use Loops In R?

In general, use loops if it improves code readability, makes it easier
to extend the code, and does not cause too big inefficiencies.  In practice, it
means use loops for repeating something slow.
Don't use loops if you can use vectorized operators instead.

For instance, look at the examples above where we were reading data
until it was good.  There is probably very little you can achieve by
avoiding loops--except making your code messy.  First, the
input will probably attempted a few times only (otherwise re-consider
how you get your data), and second, data input is probably a far
slower process than the loop overhead.  Just use the loop.

Similar arguments hold if you consider to loop over input files to read, output
files to create, figures to plot, webpage to scrape...  In all these cases
the loop overheads are negligible compared to the task performed
inside the loops.  And there are no real vectorized alternatives
anyway.

The opposite is the case where good alternatives exist.  Never do this
kind of computations in R:
```r
res <- numeric()
for(i in 1:1000) {
   res <- c(res, sqrt(i))
}
```
This loop introduces three inefficiencies.  First, and most
importantly, you are growing your results vector `res` when adding new values to it.  This involves
creating new and longer vectors again and again when the old one is
filled up.  This is very slow.  Second, as the true vectorized
alternatives exist here, one can easily speed up the code by a
magnitude or more by just switching to the vectorized code.  Third, and
this is a very important point again, this code is much harder
to read than the pure vectorized expression:
```r
res <- sqrt(1:1000)
```

We can easily demonstrate how much faster the vectorized expressions work.
Let's do this using _microbenchmark_ library (it provides nanosecond
timings).  Let's wrap these expressions in functions for easier
handling by _microbenchmark_.  We also include
a middle version where we use a loop but avoid the terribly slow
vector re-allocation by creating the right-sized result in
advance:
```{r}
seq <- 1:1000 # global variables are easier to handle with microbenchmark
baad <- function() {
   r <- numeric()
   for(i in seq)
      r <- c(r, sqrt(i))
}
soso <- function() {
   r <- numeric(length(seq))
   for(i in seq)
      r[i] <- sqrt(i)
}
good <- function()
   sqrt(seq)
microbenchmark::microbenchmark(baad(), soso(), good())
```
When comparing the median values, we can see that the middle example
is 10x slower than the vectorized example, and the first version is
almost 300x slower!

The above holds for "true" vectorized operations.  R provides many
pseudo-vectorizers (the lapply family) that may be handy in many
circumstances, but not in case you need speed.  Under the hood these
functions just use a loop.  We can demonstrate this by adding one more function to our family:
```{r}
soso2 <- function() {
   sapply(seq, sqrt)
}
microbenchmark::microbenchmark(baad(), soso(), soso2(), good())
```
As it turns out, `sapply` is much slower than plain for-loop.
Lapply-family is not a substitute for true vectorization.  However,
one may argue that `soso2` function is easier to understand than
explicit loop in `soso`.  In any case, it is easier to type when
working on the interactive console.

Finally, it should be noted that when performing truly vectorized
operations on long vectors, the overhead of R interpreter becomes
negligible and the resulting speed is almost equalt to that of the
corresponding C code.  And optimized R code can easily beat
non-optimized C.



## More about `if` and `else`

The basics of if-else blocks were covered in Chapter [Functions].
Here we discuss some more advanced aspects.

### Where To Put Else

Normally R wants to understand if a line of code belongs to an earlier
statement, or is it beginning of a fresh one.  For
instance, 
```r
if(condition) {
   ## do something
}
else {
   ## do something else
}
```
may fail.  R finishes the if-block on line 3 and thinks line 4 is
beginning of a new statement.  And gets confused by the unaccompanied
`else` as it has already forgotten
about the `if` above.  However, this code works if used inside a function, or
more generally, inside a {...} block.

But else may always stay on the same line as the the statement of the
if-block.  You are already familiar with the form
```r
if(condition) {
   ## do something
} else {
   ## do something else
}
```
but this applies more generally.  For instance:
```r
if(condition) 
   x <- 1 else {
             ## do something else
          }
```
or
```r
if(condition) { ... } else {
             ## do something else
          }
```
or
```r
if(condition) { x <- 1; y <- 2 } else z <- 3
```
the last one-liner uses the fact that we can write several statements on a
single line with `;`.  Arguably, such style is often a bad idea, but
it has it's place, for instance when creating anonymous functions
for `lapply` and friends.


### Return Value

As most commands in R, if-else block has a return value.  This is the
last value evaluated by the block, and it can be assigned to
variables.  If the condition is false, and there is no else block, `if` returns `NULL`.

Using return values of if-else statement is often a recipe for hard-to-read code but may be a good
choice in other circumstances, for instance where the multy-line form will take
too much attention away from more crucial parts of the code.  The line
```{r}
n <- 3
parity <- if(n %% 2 == 0) "even" else "odd"
parity
```
is rather easy to read.  This is the closest general construct in R corresponding to C conditional shortcuts
`parity = n % 2 ? "even" : "odd"`.

Another good place for such compressed if-else statements is 
inside anonymous functions in `lapply` and friends.  For instance, the following code
replaces `NULL`-s and `NA`-s in a list of  characters:
```{r}
emails <- list("ott@example.com", "xi@abc.net", NULL, "li@whitehouse.gov", NA)
n <- 0
sapply(emails, function(x) if(is.null(x) || is.na(x)) "-" else x)
```


## `switch`: Choosing Between Multiple Conditions

If-else is appropriate if we have a small number of
conditions, potentially only one.  Switch is a related construct that
tests a large number of equality conditions.  The best way to explain
it's syntax is through an example:
```{r}
stock <- "AMZN"
switch(stock,
       "AAPL" = "Apple Inc",
       "AMZN" = "Amazon.com Inc",
       "INTC" = "Intel Corp",
       "unknown")
```
Switch expression attempts to match the expression, here `stock = "AMZN"`, to a
list of alternatives.  If one of the alternative matches the name of
the argument, that argument is returned.  If none matches, the
nameless default argument (here "unknown") is returned.  In case there
is no default argument, `switch` returns `NULL`.

Switch allows to specify the same return value for multiple cases:
```{r}
switch(stock,
       "AAPL" = "Apple Inc",
       "AMZN" =, "INTC" = "something else",
       "unknown")
```
If the matching named argument is empty, the first non-empty named
argument is evaluated.  In this case this is "something else",
corresponding to the name "INTC".

As in case of `if`, `switch` only accepts length-1 expressions.
Switch also allows to use integers instead of character expressions,
in that case the return value list should be without names.

