# Functions

This chapter will explore how to use **functions** in R to perform advanced capabilities and actually ask questions about data. After considering a function in an abstract sense, it will discuss using built-in R functions, accessing additional functions by loading R packages, and writing your own functions.


## What are Functions?

In a broad sense, a **function** is a named sequence of instructions
(lines of code) that you may want to perform one or more times
throughout a program. They provide a way of _encapsulating_ multiple
instructions into a single "unit" that can be used in a variety of
different contexts. So rather than needing to repeatedly write down all
the individual instructions for "make a sandwich" every time you're
hungry, you can define a `make_sandwich()` function once and then just
**call** (execute) that function when you want to perform those steps.
Typically, functions also accept _inputs_ so you can make the things
slightly differently from time to time.  For instance, sometimes you may
want to call `make_sandwich(cheese)` while another time `make_sandwich(chicken)`.

In addition to grouping instructions, functions in programming languages
also allow to model the mathematical definition of function,
i.e. perform certain operations on a number of inputs that lead to an
_output_.  For instance, look at the
function `max()` that finds the largest value among numbers:

```{r, eval=FALSE}
max(1,2,3)  # 3
```
The inputs are numbers `1`, `2`, and `3` in parenthesis, usually called
**arguments** or **parameters**, and we say that these arguments are
**passed** to a function (like a football). We say that a function then
**returns** a value, number "3" in this example, which we can either
print or assign to a variable and use later.

Finally, the functions may also have **side effects**.  An example case
is the `cat()` function that just prints it's arguments.  For instance,
in case of the following line of code
```{r, eval=FALSE}
cat("The answer is", 1+1, "\n")
```
we call function `cat()` with three arguments: `"The answer is"`, `2`
(note: `1+1` will be evaluated to 2),
and `"\n"` (the line break symbol).  However, here we may not care about the return value but in the side effect&mdash;the message being printed on the screen.


## How to Use Functions

R functions are referred to by name (technically, they are values like
any other variable, just not atomic values). As in many programming
languages, we **call** a function by writing the name of the function
followed immediately (no space) by parentheses `()`. Inside the
parentheses, we put the **arguments** (inputs) to the function separated
by commas (**`,`**). Thus computer functions look just like
multi-variable mathematical functions, but usually with fancier names than `f()`.

```r
# call the sqrt() function, passing it 25 as an argument
sqrt(25)  # 5, square root of 25

# count number of characters in "Hello world" as an argument
nchar("Hello world")  # 11, note: space is a character too

# call the min() function, pass it 1, 6/8, AND 4/3 as arguments
# this is an example of a function that takes multiple args
min(1, 6/8, 4/3)  # 0.75, (6/8 is the smallest value)
```

To keep functions and ordinary variables distinct, we include empty parentheses `()` when referring to a function by name. This does not mean that the function takes no arguments, it is just a useful shorthand for indicating that something is a function.

<p class="alert alert-info">**Note:** You always need to supply the
parenthesis if you want to _call_ the function (force it to do what it
is supposed to do).  If you leave the parenthesis out, you get the
function definition printed on screen instead.  (Try this by issuing
`cat` without parenthesis on the console.)  So `cat()` is actually a _function call_ while `cat` is the function.
However, we ignore this distinction in this book.</p>

If you call any of these functions interactively, R will display the **returned value** (the output) in the console. However, the computer is not able to "read" what is written in the console&mdash;that's for humans to view! If you want the computer to be able to _use_ a returned value, you will need to give that value a name so that the computer can refer to it. That is, you need to store the returned value in a variable:

```r
# store min value in smallest.number variable
smallest_number <- min(1, 6/8, 4/3)

# we can then use the variable as normal, such as for a comparison
min_is_big <- smallest_number > 1  # FALSE

# we can also use functions directly when doing computations
phi <- .5 + sqrt(5)/2  # 1.618...

# we can even pass the result of a function as an argument to another!
# watch out for where the parentheses close!
print(min(1.5, sqrt(3)))  # prints 1.5
```

- In the last example, the resulting _value_ of the "inner" function
  (e.g., `sqrt()`) is immediately used as an argument for the middle
  function (i.e. `min()`), value of which is fed in turn to the outer
  function `print()`.  Because that value is used immediately, we don't
  have to assign it a separate variable name. It is known as an
  **anonymous variable**.
- note also that in the last example, we are solely interested in the
  side effect of the `print()` function.  It also returns it's argument
  (`min(1.5, sqrt(3))` here) but we do not store it in a variable.


## Built-in R Functions

As you have likely noticed, R comes with a variety of functions that are built into the language. In the above example, we used the `print()` function to print a value to the console, the `min()` function to find the smallest number among the arguments, and the `sqrt()` function to take the square root of a number. Here is a _very_ limited list of functions you can experiment with (or see a few more [here](http://www.statmethods.net/management/functions.html)).

<!-- TODO: Fix the formatting for this? -->

| Function Name | Description | Example |
| :------------- | :---------------------- | :------------------------------- |
| `sum(a,b,...)`  | Calculates the sum of all input values  | `sum(1, 5)` returns `6`|
| `round(x,digits)` | Rounds the first argument to the given number of digits | `round(3.1415, 3)` returns `3.142` |
| `toupper(str)`  | Returns the characters in uppercase  | `toupper("hi there")` returns `"HI THERE"` |
| `paste(a,b,...)` | _Concatenate_ (combine) characters into one value | `paste("hi", "there")` returns `"hi there"` |
| `nchar(str)` | Counts the number of characters in a string | `nchar("hi there")` returns `8` (space is a character!) |
| `c(a,b,...)` | _Concatenate_ (combine) multiple items into a _vector_ (see [chapter 7](#vectors)) | `c(1, 2)` returns `1, 2` |
| `seq(a,b)`  | Return a sequence of numbers from a to b  | `seq(1, 5)` returns `1, 2, 3, 4, 5` |


To learn more about any individual function, look them up in the R documentation by using `?FunctionName` account as described in the previous chapter.

<!-- TODO: Make this a link to the chapter?? Why won't it work? -->

<p class="alert alert-info">"Knowing" how to program in a language is to some extent simply "knowing" what provided functions are available in that language. Thus you should look around and become familiar with these functions... but **do not** feel that you need to memorize them! It's enough to simply be aware "oh yeah, there was a function that sums up numbers", and then be able to look up the name and argument for that function.</p>


## Loading Functions

Although R comes with lots of built-in functions, you can always use
more! **Packages** (also known as **libraries**) are additional sets of
R functions (and data and variables) that are written and published by the R community. Because
many R users encounter the same data management/analysis challenges,
programmers are able to use these libraries and thus benefit from the
work of others (this is the amazing thing about the open-source
community&mdash;people solve problems and then make those solutions
available to others).  Popular packages include _dplyr_ for
manipulating data, _ggplot2_ for visualizations, and _data.table_ for
handling large datasets.

Most of the R packages **do not** ship with the R
software by default, and need to be downloaded (once) and then loaded
into your interpreter's environment (each time you wish to use
them). While this may seem cumbersome, it is a necessary trade-off
between speed and size.  R software would be huge and slow if it would
include _all_ available packages.

Luckily, it is quite simple to install and load R packages from within
R. To do so, you'll need to use the _built-in_ R functions
`install.packages` and `library`. Below is an example of installing and
loading the `stringr` package (which contains many handy functions for working with character strings):

```r
# Install the `stringr` package. Only needs to be done once on your machine
install.packages("stringr")
```
<p class="alert alert-warning">
We stress here that you need to install each package only once per computer.
As installation may be slow and and resource-demanding, you **should not
do it repeatedly inside your script!**.  Even more, if your script is
also run by other users on their computers, you should get their
explicit consent before installing additional software for them!
The easiest remedy is to solely rely on manual installation.
</p>

Exactly the same syntax&mdash;`install.packages("stringr")`&mdash;is
also used for re-installing it.  You may want to re-install it if a
newer version comes out, of if you upgrade your R and receive warnings 
about the package being built under a previous version of R.
  
After installation, the easiest way to get access to the functions is by
_loading the package_:
```{r, eval=FALSE}
# Load the package (make stringr() functions available in this R session/program)
library("stringr") # quotes optional here
```
This makes all functions in the _stringr_ package available for R (see [the documentation](https://cran.r-project.org/web/packages/stringr/stringr.pdf) for a list of functions included with the `stringr` library).  For
instance, if we want to pad the word "justice" from left with tildes to create a
width-10 string, we can do
```{r, eval=FALSE}
str_pad("justice", 10, "left", "~")  # "~~~justice"
```
We can use `str_pad` function without any additional hassle because `library()` command made it
available.

This is an easy and popular approach.  However, what happens if more than one package call a function by the same name?  For
instance, many packages implement function `filter()`.  In this case
the more recent package will _mask_ the function as defined by the
previous package.  You will also see related warnings when you load the
library.  In case you want to use a masked
function you can write something like `package::function()` in order to
call it.  For instance, we can do the example above with
```{r, eval=FALSE}
stringr::str_pad("justice", 10, "left", "~")  # "~~~justice"
```
This approach&mdash;specifying _namespace_ in front of the
function&mdash;ensures we access the function in the right package.  If
we call all functions in this way, we don't even need to load the
package with `library()` command.  This is the preferred approach if you
only need few functions from a large library.



## Writing Functions
Even more exciting than loading other peoples' functions is writing your own. Any time that you have a task that you may repeat throughout a script&mdash;or you simply want to organize your thinking&mdash;it's good practice to write a function to perform that task. This will limit repetition and reduce the likelihood of errors... as well as make things easier to read and understand (and thus identify flaws in your analysis).

Functions are named like any other variable, so we use the _assignment operator_ (**`<-`**) to store a new `function` in a variable. It is [best practice ](http://style.tidyverse.org/functions.html) to assign functions names as verbs, also written in **snake_case** (just like other variables).
The best way to understand the syntax for defining a function is to look at an example:

```r
# A function named `make_full_name` that takes two arguments
# and returns the "full name" made from them
make_full_name <- function(first_name, last_name) {
  # Function body: perform tasks in here
  full_name <- paste(first_name, last_name)

  # Return: what you want the function to output
  return(full_name)
}

# Call the `make_full_name` function with the values "Alice" and "Kim"
my.name <- make_full_name("Alice", "Kim")  # "Alice Kim"
```

Functions have a couple of pieces to them:

- **Arguments**: the value assigned to the function variable uses the syntax `function(...)` to indicate that you are creating a function (as opposed to a number or character string). The values put between the parentheses are variables that _will contain_ the values passed in as **arguments**. For example, when we call `make_full_name("Alice", "Kim")`, the value of the first argument (`"Alice"`) will be assigned to the first variable (`first.name`), and the value of the second argument (`"Kim"`) will be assigned to the second variable (`last.name`).

    Importantly, we could have made the argument names anything we wanted (`name.first`, `given.name`, etc.), just as long as we then use _that variable name_ to refer to the argument while inside the function. Moreover, these argument variable names _only apply_ while inside the function. You can think of them like "nicknames" for the values. The variables `first.name`, `last.name`, and `full.name` only exist within this particular function.

- **Body**: The body of the function is a **block** of code that falls between curly braces **`{}`** (a "block" is represented by curly braces surrounding code statements). Note that cleanest style is to put the opening `{` immediately after the arguments list, and the closing `}` on its own line.

    The function body specifies all the instructions (lines of code) that your function will perform. A function can contain as many lines of code as you want&mdash;you'll usually want more than 1 to make it worth while, but if you have more than 20 you might want to break it up into separate functions. You can use the argument variables in here, create new variables, call other functions... basically any code that you would write outside of a function can be written inside of one as well!

- **Return value**: You can specify what output a function produces by calling the `return()` function and passing that the value that you wish _your function_ to return (output). The `return()` function will execute instructions that end the current function and _return_ the flow of code execution to wherever this function was called from. Note that even though we returned a variable called `full_name`, that variable was _local_ to the function and so doesn't exist outside of it; thus we have to take the returned value and assign it to a new variable (as with `name <- make_full_name("Alice", "Kim")`).

    Because the `return()` call exits the function, it is usually the last line of code in the function. Technically, the function will return the last element defined in your function, so it is unnecessary to write a _return_ statement.

We can call (execute) a function we defined the same way we called built-in functions. When we do so, R will take the **arguments** we passed in (e.g., `"Alice"` and `"Kim"`) and assign them to the _argument variables_. Then it executes each line of code in the **function body** one at a time. When it gets to the `return()` call, it will end the function and return the given value, which can then be assigned to a different variable outside of the functions.


## Conditional Statements
Functions are a way to organize and control the flow of execution (e.g., what lines of code get run in what order). In R, as in other languages, we have one other way of controlling program flow, and that is by specifying different instructions that can be run based on a different set of conditions. **Conditional statements** allow us to specify different chunks of code to run when given different contexts, which is often valuable within functions.

In an abstract sense, an conditional statement is saying:

```
IF something is true
  do some lines of code
OTHERWISE
  do some other lines of code
```

In R, we write these conditional statements using the keywords **`if`** and **`else`** and the following syntax:

```r
if (condition) {
  # lines of code to run if condition is TRUE
} else {
  # lines of code to run if condition is FALSE
}
```

(Note that the the `else` needs to be on the same line as the closing `}` of the `if` block. It is also possible to omit the `else` and its block).

The `condition` can be any variable or expression that resolves to a logical value (`TRUE` or `FALSE`). Thus both of the below conditional statements are valid:

```r
porridge_temp <- 115 # in degrees F
if (porridge_temp > 120) {
  print("This porridge is too hot!")
}

too_cold <- porridge_temp < 70
# a logical value
if (too_cold) {
  print("This porridge is too cold!")
}
```

Note, we can extend the set of conditions evaluated using an `else if` statement. For example:

```r
# Function to determine if you should eat porridge
test_food_temp <- function(temp) {
  if (temp > 120) {
    status <- "This porridge is too hot!"
  } else if (temp < 70) {
    status <- "This porridge is too cold!"
  } else {
    status <- "This porridge is just right!"
  }
  return(status)
}
# Use function on different temperatures
test_food_temp(119)  # "This porridge is just right!"
test_food_temp(60)   # "This porridge is too cold!"
test_food_temp(150)  # "This porridge is too hot!"
```

See more about `if` and `else` in [Appendix C](#control-structures).

## Resources {-}
- [R Function Cheatsheet](https://cran.r-project.org/doc/contrib/Short-refcard.pdf)
- [User Defined R Functions](http://www.statmethods.net/management/userfunctions.html)
